---
title: "Tutorial: countsplitting and scran" 
output: rmarkdown::html_vignette
bibliography: latent.bib
vignette: >
  %\VignetteIndexEntry{"Scran Tutorial: countsplitting and scran"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Before using this tutorial, we recommend that you read through our ``introduction to count splitting" tutorial to understand our method in a simple package with simulated data. 

In this tutorial, we reproduce the analysis from the [scran tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/scran/inst/doc/scran.html). The steps of the analysis include preprocessing, graph-based clustering, and identifying differentially expressed marker genes for the clusters. 

# Install scran

If you don't already have ``scran``, you will need to run:

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scran")
```

Next, you should load the package, along with others that we will use in this tutorial.

```{r, message=FALSE, warning=FALSE}
library(scran)
library(tidyverse)
library(countsplit)
```

# 

