---
title: "Tutorial: differential expression analysis on single cell RNA-seq data" 
output: rmarkdown::html_vignette
bibliography: latent.bib
vignette: >
  %\VignetteIndexEntry{"Tutorial: differential expression analysis on single cell RNA-seq data"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

In this tutorial, we use two simple simulated data sets to demonstrate how to use the countsplit package and base R functions (e.g. glm, kmeans) to cluster and test for differential expression in scRNA-seq data. See our other tutorials on this website to learn how to integrate the countsplit package into three existing scRNA-seq data analysis workflows in R: `Seurat`, `scran`, and `monocle3`.
.
We start by loading the packages we will be working with. Make sure that ``remotes`` is installed by running ``install.packages("remotes")``, then type

```{r,eval=FALSE}
remotes::install_github("anna-neufeld/countsplit")
```

```{r, message=FALSE}
library(countsplit)
library(ggplot2)
library(patchwork)
```

# Simulated data with no true signal. 

First suppose that we have $n=1000$ cells and $p=200$ genes. Suppose that every count $\textbf{X}_{ij}$ is drawn from a $\text{Poisson}(5)$ distribution. We first generate this data. 

```{r}
set.seed(1)
n <- 1000
p <- 200
X <- matrix(rpois(n*p, lambda=5), nrow=n)
```

In this tutorial, we are interested in knowing which genes are differentially expressed across discrete cell types. We will cluster this data to estimate cell types and then test for differential expression. Since there are no true cell types and no truly differentially expressed genes in this data, a test of differential expression that controls the Type 1 error rate should give uniformly distributed p-values on this data.

## The naive method

First, we will demonstrate that estimating the clusters and testing for differential expression using the same data does not control the Type 1 error rate. We refer to the practice of using the same data for clustering and differential expression testing as the "naive method" or "double dipping". 

```{r}
clusters.full <- kmeans(log(X+1), centers=2)$cluster
results.naive <- t(apply(X, 2, function(u) summary(glm(u~as.factor(clusters.full), family="poisson"))$coefficients[2,]))
head(results.naive)
```

The first line clusters the cells by applying k-means clustering with k=2 on the log-transformed data with a pseudocount of 1, and saves the cluster assignments as `clusters.full`. The second line tests for differential gene expression using Poisson GLMs. For every gene $X_j$ in X, we fit a Poisson GLM of $X_j$ on `clusters.full`, and save the summary of the slope coefficient in `results.naive`. As shown in the output, we have saved a slope coefficient estimate, a standard error, a z-value, and a p-value for every gene in the dataset. 

Even in these first 6 rows of results, we can see that the naive method assigns small p-values to many genes, despite the fact that no genes are truly differentially expressed in this data. We can make a uniform QQ-plot of the p-values for the naive method to see that they are not uniformly distributed and thus do not control the Type 1 error. 

```{r}
ggplot(data=NULL, aes(sample=results.naive[,4]))+geom_qq(distribution="qunif")+geom_abline(col="red")
```

## Count splitting 

We now address the issue using count splitting. The key steps are (1) running the ``countsplit`` function to get ``Xtrain`` and ``Xtest`` and then (2) clustering the data using `Xtrain`, and then fitting a GLM to test for association between those clusters and each column of `Xtest`. The ``countsplit`` function returns a list, which we call ``split`` here, which contains the training set and the test set.

The parameter `epsilon`, which must fall between 0 and 1, governs the relative amount of information from `X` that goes into the training set. When `epsilon` is very close to 0, the training matrix will be extremely sparse and the test matrix will look a lot like `X`. When epsilon is very close to 1, the training matrix will be nearly identical to `X`, but the test matrix will be extremely sparse. The default in the `countsplit` function is to set epsilon=0.5. See [our preprint](https://arxiv.org/abs/2207.00554) for more information.  

```{r}
set.seed(2)
split <- countsplit(X, epsilon=0.5)
names(split)
Xtrain <- split$train
Xtest <- split$test
```

Like before, we will cluster using k-means with log-transformed data and we will fit Poisson GLMs for differential expression. Unlike before, we will run the clustering on `Xtrain` and use `Xtest` as the response in our GLMs.

```{r}
clusters.train <- kmeans(log(Xtrain+1), centers=2)$cluster
results.countsplit <- t(apply(Xtest, 2, function(u) summary(glm(u~as.factor(clusters.train), family="poisson"))$coefficients[2,]))
head(results.countsplit)
```

We can see from the summary output that the p-values for the first 6 genes are much larger. When we make the same uniform QQ-plot as before, we see that the p-values obtained from count splitting are uniformly distributed. Since no genes are differentially expressed in our data, this means that count splitting controls the Type 1 error.


```{r}
ggplot(data=NULL, aes(sample=results.countsplit[,4]))+geom_qq(distribution="qunif")+geom_abline(col="red")
```

In summary, count splitting controls the Type 1 error when there is no true signal in the data. 

# Simulated data with true signal

We now demonstrate the performance of count splitting on a simple simulated dataset that contains two true clusters. We first randomly assign the cells to one of two true clusters. We then generate data such that $X_{ij} \sim \mathrm{Poisson}(\Lambda_{ij})$. Genes 1-10 are differentially expressed-- for $j=1,\ldots,10$, $\Lambda_{ij} = 5$ for cells in cluster $0$ and $\Lambda_{ij}=10$ for cells in cluster $1$. Genes 11-200 are not differentially expressed ($\Lambda_{ij}=5$ for all cells). 

```{r}
set.seed(1)
n <- 1000
p <- 200
clusters.true <- rbinom(n, size=1, prob=0.5)
Lambda <- matrix(5, nrow=n, ncol=p)
Lambda[clusters.true==1, 1:10] <- 10
X <-apply(Lambda,1:2,rpois,n=1)
```

We now count split the data and save `Xtrain` and `Xtest` for later use.

```{r}
split <- countsplit(X, epsilon=0.5)
Xtrain <- split$train
Xtest <- split$test
```

## Effect of using Xtrain for cluster estimation. 

First, let's look at the effect of count splitting on our ability to estimate the true clusters. 
If we use all of our data `X` to estimate the clusters, we make only 5 errors. 

```{r}
clusters.full <- kmeans(log(X+1), centers=2)$cluster
table(clusters.true, clusters.full)
```

If instead we use only `Xtrain` to estimate the clusters, we make a few additional errors, but we still come very close to estimating the true clusters.

```{r}
clusters.train <- kmeans(log(Xtrain+1), centers=2)$cluster
table(clusters.true, clusters.train)
```

The fact that cluster 0 in the true clustering maps to cluster 2 in the training set clustering is unimportant, but we will remember this later when we attempt to compare models. 


## Effect of using Xtest for inference

We now compare what happens when we regress $X_j$ on the true clusters to what happens if we regress $X_j^{\mathrm{test}}$ on the true clusters. We use the code below to fit a Poisson GLM of every gene on the true clusters; we save both the slope and the intercept. We are interested in the GLM slopes, as the GLM slopes are non-zero if the gene is differentially expressed across the clusters. 

```{r}
coeffs.X <- t(apply(X, 2, function(u) summary(glm(u~as.factor(clusters.true), family="poisson"))$coefficients[,1]))
coeffs.Xtest <- t(apply(Xtest, 2, function(u) summary(glm(u~as.factor(clusters.true), family="poisson"))$coefficients[,1]))
```

The plot below shows that the slopes resulting from `X` and the slopes resulting from `Xtest` tend to fall on the diagonal line `y=x` and so tend to be approximately equal to eachother. The intercepts, on the other hand, get shifted by `log(0.5)` when we use the `Xtest` rather than `X`. This is as we would expect (see Section 4.1 of [our preprint](https://arxiv.org/abs/2207.00554)). 

```{r}
differentially_expressed = as.factor(c(rep(1,10), rep(0,190)))
p1 <- ggplot(data=NULL, aes(x=coeffs.X[,1], y=coeffs.Xtest[,1], col=differentially_expressed))+
  geom_point()+
  geom_abline(intercept= log(0.5), slope=1, col="red")+
  geom_abline(intercept= 0, slope=1, col="red", lty=2)+
  coord_fixed()+xlim(0,2)+ylim(0,2)+
  xlab("Intercepts from X")+ ylab("Intercepts from Xtest")+
  ggtitle("Intercepts")+theme_bw()
p2 <- ggplot(data=NULL, aes(x=coeffs.X[,2], y=coeffs.Xtest[,2], col=differentially_expressed))+
  geom_point()+
  geom_abline(intercept=0, slope=1, col="red")+
  coord_fixed()+xlim(-0.15,1)+ylim(-0.15,1)+
  xlab("Slopes from X")+ ylab("Slopes from Xtest")+
  ggtitle("Slopes")+theme_bw()
p1 + p2 + plot_layout(guides="collect")
```

## Overall comparison of count splitting to ideal method

Finally, we compare the overall slope and intercept estimates that we get from count splitting to what we would get in the ideal setting where we get to regress $X_j$ on the true clusters. Note that `coeffs.ideal` is identical to `coeffs.X` from the previous section; we reproduce it here for convenience. Note also that we regress `Xtest` on `as.factor(clusters.train==1)`, rather than `as.factor(clusters.train)`, simply because we saw above that the "cluster 1" in the training set maps to the "cluster 1" in the true clustering, and this consistency will ensure that the coefficients from the two models have the same sign. 

```{r}
coeffs.ideal <- t(apply(X, 2, function(u) summary(glm(u~as.factor(clusters.true), family="poisson"))$coefficients[,1]))
coeffs.countsplit <- t(apply(Xtest, 2, function(u) summary(glm(u~as.factor(clusters.train==1), family="poisson"))$coefficients[,1]))
p1 <- ggplot(data=NULL, aes(x=coeffs.ideal[,1], y=coeffs.countsplit[,1], col=differentially_expressed))+
  geom_point()+
  geom_abline(intercept= log(0.5), slope=1, col="red")+
  geom_abline(intercept= 0, slope=1, col="red", lty=2)+
  coord_fixed()+xlim(0,2)+ylim(0,2)+
  xlab("Intercepts from ideal method")+ ylab("Intercepts from count splitting")+
  ggtitle("Intercepts")
p2 <- ggplot(data=NULL, aes(x=coeffs.ideal[,2], y=coeffs.countsplit[,2], col=differentially_expressed))+
  geom_point()+
  geom_abline(intercept=0, slope=1, col="red")+
  coord_fixed()+xlim(-0.15,1)+ylim(-0.15,1)+
  xlab("Slopes from ideal method")+ ylab("Slopes from count splitting")+
  ggtitle("Slopes")
p1 + p2 + plot_layout(guides="collect") & theme_bw()
```

Overall, we see general agreement between the parameters estimated via countsplitting and those estimated via the ideal method. The slopes tend to be the same, whereas the intercepts are shifted by `log(0.5)`, as expected based on [our preprint](https://arxiv.org/abs/2207.00554). The slopes are the quantities that we care to do inference on, as they measure differential expression across clusters. 





